From clusters to queries: exploiting uncertainty in the modularity landscape of complex networks

### 2018-06-25 article/article.tex
Removed faulty results
Trying to fix bug with HPC. Code functions on this pc but not hpc...

\begin{figure}[t]
    \includegraphics[width=0.48\textwidth]{images/search_space_3.png}
    \caption{Modularity search space of an \textit{E. coli} metabolic network.
 Distance between partitions is calculated using variation of information \cite{meilua2003comparing} and dimensionality reduction is performed using curvilinear component analysis \cite{demartines1997curvilinear}.
  The inset (top) demonstrates the landscape of the high modularity region.
  Figure generated with the software of Good \textit{et al.} \cite{good2010performance}.}
    \label{fig:modular_search_space}
\end{figure}

@@

\section{Results}
\subsection{Cross-validation method}
\label{sec:cross_validation}
In this study we test a small number of labels that we intend to use in order to evaluate how well our method correctly generalises to discover unlabelled vertices.
We would like to test a significantly smaller number of seed nodes than two class classification methods used in previous studies, which use leave-one-out cross validation \cite{kohler2008walking}.
The cross validation procedure we devise is described as follows and depends on the size of the community and the number of initial seed labels being used.
For this work we would like to capture binary classification  performance, \textit{true positives (tp)}, \textit{true negatives (tn), false positives (fp)} and \textit{false negatives (fn)}, on our datasets of community memberships.
In order to generate the different sets for the cross validation we take each label set and generate unique sample sets of vertices from the known true positive labels\footnote{Note that although each node can have multiple labels, in the tests performed in this study we only consider performance on individual labels as this is likely the situation one would encounter with limited label sets in practice.}.

As the seed label sets can be as small as 3 vertices, exhaustive cross validation was not possible for all labelling schemes.
Consequently, cross validation is either conducted on an exhaustive set of all possible $\binom{|S|}{s}$ unique labellings or 120 seed queries sampled randomly without replacement from the possible subsets \footnote{This is equivalent to $\binom{10}{3}$ combinations, an exhaustive sample would not be possible for the larger communities in this study.}, where $S$ is the set of gold standard true labels and $s$ is the size of the randomly selected seed sets.
It should therefore be noted that presented receiver operator characteristic (ROC) scores are dependent on community sizes.
We also note that we do not consider separate training and test sets in this study as the method does not use any examples when building the index space from partitions of the graph.

Formally, the steps for this procedure with a given label set $S$ are outlined as follows:
\begin{itemize}
 \item Generate up to 120 unique subsets of size $s$, randomly sampled without replacement (set $T$).
 \item For each test subset $\{\forall S_t \in T | S_t \subset S \wedge |S_t| = s \}$, generate the $\mu_i(S_t)$ score for all vertices in the network
 \item Exclude vertices in $S_t$ from the test
 \item Consider all nodes in $S$ but not in $S_t$ to be the true positive set.
 \item Consider each of the nodes in $V$ not in $S$ to be the true negative set.
 \item Generate an average ROC curve interpolated from all test subsets.
 \end{itemize}

@@

%The networks and resulting area under the curve (AUC) of ROC statistics are shown in Table \ref{table:dataset_table}.
ROC curves generated with the cross validation procedure described in Section \ref{sec:cross_validation} are shown in Figure \ref{fig:quality_label_rocs}.
These results represent ROC curves and mean AUC statistics.
For all the datasets in this study, binary community classification is high, with an average AUC scores above 0.87, even when using only a single seed node.
In the case of the \textit{Arabidopsis thaliana} network the protein complexes tested are significantly smaller than for other networks and so a comparison of seed sizes is not possible.
However, in the other networks studied, using 3 or more seed nodes appears to improve results, though for 15 seed nodes the results are not significantly better than with 7 seeds.
The results of our method are comparable to the random walk with restart approach, which performs better on the datasets tested with the exception of the EU email dataset for which our model performs significantly better.

@@

\begin{figure}[t]
\includegraphics[width=0.48\textwidth]{images/ppi_example_combined.eps}
\caption{An example of the GO term \textit{``SNAP receptor activity''} (16 nodes) on the \textit{Arabidopsis thaliana} protein-protein interaction network.
(A) Shows the full network with 3 randomly chosen query nodes from the full query set.
(B) Shows a reduced view of the network where a threshold of $\mu_i(S) > 0.8$ is selected, reducing the network to 404 nodes.
Blue nodes indicates the 3 randomly selected query nodes, yellow the remaining nodes in the term and the shade of red relates to the strength of the relationship each node has to the query set.
(C) Shows an example ROC curve from the cross validation described in the main text.
}
\label{fig:query_example}
\end{figure}

In a practical situation the high quality label sets described above are not likely to be available.
Consequently, we wish to highlight that the method also serves as a useful network reduction tool where the quality of labels is not well defined.
In order to achieve this goal, for each of the biological data sets in this study, we acquired all Gene Ontology terms for each of the proteins from AmiGO \cite{carbon2008amigo}.
Gene Ontology is a controlled vocabulary of terms associated with biological functions in three broad categories of Cellular Components (CC),  Metabolic Functions (MF) and Biological Processes (BP), forming a hierarchy of terms and associated sub terms.
The use of a $\mu_i(S)$ score on a network is visually demonstrated for the \textit{Arabidopsis PPI} dataset shown in Figure \ref{fig:query_example} with the hand selected GO term \textit{``SNAP receptor activity''}.

@@

The fact that the method is able to uncover small protein complexes seems to contradict the principle that modularity maximisation algorithms have a resolution limit \cite{fortunato2007resolution}.
Whilst this appears to be the case it is important to note that the resolution limit applies to a single partition of space.
Further work is needed to investigate why small communities are still detectable.
However, we speculate that it is likely due to the fact that the co-classification of vertices between different partitions remains fairly tolerant to changes.
In other words, whilst the large communities they reside in changes significantly, the small cluster of nodes is always clustered in the same community, regardless of the partition.
We do note that, where communities are very small, any approach will be extremely sensitive to false positive and false negative results.
As such, this should be considered when using any method of this form as an exploratory tool.

The results also appear to be highly tolerant to a small number of seed nodes.
This is interesting as in most sampled cases the relevant nodes are unlikely to be direct neighbours.
From the perspective of exploratory studies, this implies that a small number of query vertices can be used to find potentially related vertices.

@@

\section{Conclusions}
This paper has presented a novel approach to semi-supervised community detection utilising a consensus of high scoring partitions computed with the popular modularity maximisation approach.
Previously the glassy search space of this optimisation algorithm has been seen as a major limitation.
However, in this work we consider each locally optimal partition to be information regarding the true multi-class labels that are likely present in real networks.
The approach presented here differs from other ensemble approaches in that the objective is to provide a probabilistic framework for label classification.
Performance was shown to be strong on both synthetically generated networks and real-world ground truth communities with relatively small sets of labels.
In the case of synthetic networks communities are correctly detected up to the detectability threshold.
For real world networks with small label sets, average AUC scores were above 0.87 and comparable to the random walk with restart method for the high quality datasets tested in this study.

>>>>

\begin{figure}[t]
    \includegraphics[width=0.48\textwidth]{images/e_coli_search_space.eps}
    \caption{Modularity search space of an \textit{E. coli} metabolic network.
 Distance between partitions is calculated using variation of information \cite{meilua2003comparing} and dimensionality reduction is performed using curvilinear component analysis \cite{demartines1997curvilinear}.
  The inset (top) demonstrates the landscape of the high modularity region.
  Figure generated with the software of Good \textit{et al.} \cite{good2010performance}.}
    \label{fig:modular_search_space}
\end{figure}

@@

\section{Results}
\subsection{Cross-validation method}
\label{sec:cross_validation}
In this study we test a small number of labels that we intend to use in order to evaluate how well our method correctly generalises to discover unlabelled vertices.
We would like to test a significantly smaller number of seed nodes than two class classification methods used in previous studies, which use leave-one-out cross validation \cite{kohler2008walking}.
The cross validation procedure we devise is described as follows and depends on the size of the community and the number of initial seed labels being used.

For this work we would like to capture binary classification  performance, \textit{true positives (tp)}, \textit{true negatives (tn), false positives (fp)} and \textit{false negatives (fn)}, on our datasets of community memberships.
In order to generate the different sets for the cross validation we take each label set and generate unique sample sets of vertices from the known true positive labels.

As the seed label sets can be as small as 3 vertices, exhaustive cross validation was not possible for all labelling schemes.
Consequently, cross validation is either conducted on an exhaustive set of all possible $\binom{|S|}{s}$ unique labellings or 120 seed queries sampled randomly without replacement from the possible subsets \footnote{This is equivalent to $\binom{10}{3}$ combinations, an exhaustive sample would not be possible for the larger communities in this study.}, where $S$ is the set of gold standard true labels and $s$ is the size of the randomly selected seed sets.

As the selected seed sets can be contained within multiple communities, we consider the set of true positives not to be the community for which the seed set is randomly selected, but all communities for which that seed set is a subset of.
This is because the purpose of the approach presented within this paper is to distinguish between different communities in a context specific manner, if the overlap between two communities is represented by the seed nodes this should be considered in the tests.

It should therefore be noted that presented receiver operator characteristic (ROC) scores are dependent on community sizes.
We also note that we do not consider separate training and test sets in this study as the method does not use any examples when building the index space from partitions of the graph.

Formally, the steps for this procedure with a given label set $S$ are outlined as follows:
\begin{itemize}
 \item Generate up to 120 unique subsets of size $s$, randomly sampled without replacement (set $T$).
 \item For each test subset $\{\forall S_t \in T | S_t \subset S \wedge |S_t| = s \}$, generate the $\mu_i(S_t)$ score for all vertices in the network
 \item Exclude vertices in $S_t$ from the test
 \item Consider the true community membership of $S_t$ to be the true positive set.
 \item Consider each of the nodes in $V$ not in $S$ to be the true negative set.
 \item Generate an average ROC curve interpolated from all test subsets.
 \end{itemize}

@@

%The networks and resulting area under the curve (AUC) of ROC statistics are shown in Table \ref{table:dataset_table}.
ROC curves generated with the cross validation procedure described in Section \ref{sec:cross_validation} are shown in Figure \ref{fig:quality_label_rocs}.
These results represent ROC curves and mean AUC statistics.

The results of our method are comparable to the random walk with restart approach, which performs better on the datasets tested with the exception of the EU email dataset for which our model produces higher AUC scores.

In the case of the \textit{Arabidopsis thaliana} network the protein complexes tested are significantly smaller than for other networks and so a comparison of seed sizes is not possible.
However, in the other networks studied, using 3 or more seed nodes appears to improve results, though for 15 seed nodes the results are not significantly better than with 7 seeds.
The lower quality labels within the Arabidopsis protein complex dataset likely explains the significant difference in results when compared with other protein interaction datasets.

@@

% \begin{figure}[t]
% \includegraphics[width=0.48\textwidth]{images/ppi_example_combined.eps}
% \caption{An example of the GO term \textit{``SNAP receptor activity''} (16 nodes) on the \textit{Arabidopsis thaliana} protein-protein interaction network.
% (A) Shows the full network with 3 randomly chosen query nodes from the full query set.
% (B) Shows a reduced view of the network where a threshold of $\mu_i(S) > 0.8$ is selected, reducing the network to 404 nodes.
% Blue nodes indicates the 3 randomly selected query nodes, yellow the remaining nodes in the term and the shade of red relates to the strength of the relationship each node has to the query set.
% (C) Shows an example ROC curve from the cross validation described in the main text.
% }
% \label{fig:query_example}
% \end{figure}

In a practical situation the higher quality label sets described above are not likely to be available.
Consequently, we wish to highlight that the method also serves as a useful network reduction tool where the quality of labels is not well defined.
In order to achieve this goal, for each of the biological data sets in this study, we acquired all Gene Ontology terms for each of the proteins from AmiGO \cite{carbon2008amigo}.
Gene Ontology is a controlled vocabulary of terms associated with biological functions in three broad categories of Cellular Components (CC),  Metabolic Functions (MF) and Biological Processes (BP), forming a hierarchy of terms and associated sub terms.

@@

The fact that the method is able to uncover small protein complexes seems to contradict the principle that modularity maximisation algorithms have a resolution limit \cite{fortunato2007resolution}.
Whilst this appears to be the case it is important to note that the resolution limit applies to a single partition of space.
Further work is needed to investigate why small communities are still detectable.
However, we speculate that it is likely due to the fact that the co-classification of vertices between different partitions remains fairly tolerant to changes.
In other words, the small cluster of nodes is always clustered in the same community, regardless of the partition.
We do note that, where communities are very small, any approach will be extremely sensitive to false positive and false negative results.
As such, this should be considered when using any method of this form as an exploratory tool.

The approach also appears to be tolerant to a small number of seed nodes.
This is interesting as in most sampled cases the relevant nodes are unlikely to be direct neighbours.
From the perspective of exploratory studies, this implies that a small number of query vertices can be used to find potentially related vertices.

@@

\section{Conclusions}
This paper has presented a novel approach to semi-supervised community detection utilising a consensus of high scoring partitions computed with the popular modularity maximisation approach.
Previously the glassy search space of this optimisation algorithm has been seen as a major limitation.
However, in this work we consider each locally optimal partition to be information regarding the true multi-class labels that are likely present in real networks.
The approach presented here differs from other ensemble approaches in that the objective is to provide a probabilistic framework for label classification.
Performance was shown to be strong on both synthetically generated networks and real-world ground truth communities with relatively small sets of labels.
In the case of synthetic networks communities are correctly detected up to the detectability threshold.
For real world networks with small label sets, average AUC scores were comparable to the random walk with restart method for the high quality datasets tested in this study.
